A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-08-11 appendix.tex
msre chapter edits

\section{Numerical Implementation} \label{sec:implementation}

In Section \ref{sec:theory}, I presented the theoretical basis for the hybrid
$S_N$-diffusion method and demonstrated the method on a simple problem. Now I will discuss
numerical implementation details of group
constant data processing and the neutron diffusion, $S_N$ neutron transport, and hybrid
$S_N$-diffusion solvers. I implemented all numerical solvers and analysis scripts in the Python
programming language. First, I will discuss the material group constant data generation and
postprocessing steps. Next, I present the implementation details of the neutron
diffusion and $S_N$ numerical methods. Lastly, I present how they are coupled to form the hybrid
$S_N$-diffusion method.

\subsection{Group Constant Data Generation}

The group constants required by either or both neutron diffusion and $S_N$ neutron transport
methods are:
%
\begin{itemize}
  \item $\Sigma_{t,g}$: Macroscopic total cross section in group $g$,
  \item $\Sigma_{r,g}$: Macroscopic removal cross section in group $g$,
  \item $\Sigma_s^{g'\rightarrow g}$: Macroscopic group-to-group scattering cross section matrix,
  \item $\Sigma_{s,l}^{g'\rightarrow g}$: $l$-th Legendre moment of the macroscopic
    group-to-group scattering cross section matrix,
  \item $\Sigma_{sp,l}^{g'\rightarrow g}$: $l$-th Legendre moment of the macroscopic
    group-to-group scattering production cross section matrix,
  \item $D_g$: $P_1$-based diffusion coefficient in group $g$,
  \item $\nu\Sigma_{f,g}$: Product of the average number of neutrons produced per fission and the
    macroscopic fission cross section in group $g$,
  \item $\chi_g$: Neutron fission spectrum in group $g$.
\end{itemize}
%
These group constants are generated using OpenMC's multigroup cross section generation capability
\cite{boyd_multigroup_2019} and postprocessed using a Python script into JSON format files.
$\Sigma_{r,g}$ is the only quantity that OpenMC does not directly provide. It is calculated as
%
\begin{align}
  \Sigma_{r,g} =& \sum^G_{g'\neq g}\Sigma_s^{g\rightarrow g'}+\Sigma_{a,g}-\left(\Sigma_{sp}^{g
    \rightarrow g} - \Sigma_s^{g\rightarrow g}\right)
  \shortintertext{where}
      \Sigma_{a,g} =& \mbox{ macroscopic absorption cross section in group $g$,} \nonumber \\
      \Sigma_{sp}^{g\rightarrow g} =& \mbox{ macroscopic scattering production cross section from
      group $g$ to $g$.} \nonumber
\end{align}
%
$\Sigma_{r,g}$ primarily represents the loss of neutrons from group $g$ through outscattering and
absorption. $\Sigma_{sp}^{g\rightarrow g}$ incorporates neutron multiplication effects from neutron
knockout reactions into the scattering cross section. Neutron knockout reactions are commonly
tallied as scattering reactions, and $\Sigma_{r,g}$ is a convenient term to
incorporate neutron knockout effects into the neutron diffusion equations. I ran all test cases for
$S_N$ and hybrid method calculations with up to the 2nd Legendre moments of the scattering
cross sections ($L=2$).

OpenMC uses the $P_1$ flux-limited formulation \cite{pomraning_flux-limited_1984} for calculating
$D_g$ as follows
%
\begin{align}
  D_g =& \frac{1}{3\Sigma_{tr,g}},
  \shortintertext{where}
  \Sigma_{tr,g} =& \frac{\langle\Sigma_{t,g}\phi_g\rangle-\langle\Sigma_{s1,g}\phi_g\rangle}
  {\langle\phi_g\rangle}, \nonumber \\
  \langle\Sigma_{t,g}\phi_g\rangle =& \int_{r\in V}dr \int_{4\pi}d\Omega\int^{E_{g-1}}_{E_g}dE\
  \Sigma_{t,g}(r,E)\Psi(r,E,\Omega), \nonumber \\
  \langle\Sigma_{s1,g}\phi_g\rangle =& \int_{r\in V}dr \int_{4\pi}d\Omega\int^{E_{g-1}}_{E_g}dE
  \int_{4\pi}d\Omega'\int^{\infty}_0dE'\int^1_{-1}d\mu\ \mu\Sigma_s(r,E'\rightarrow E,\Omega'\cdot
  \Omega)\phi(r,E',\Omega'), \nonumber \\
  \langle \phi \rangle =& \int_{r\in V}dr\int_{4\pi}d\Omega\int^{E_{g-1}}_{E_g}dE\ \Psi(r,E,\Omega)
  .\nonumber
\end{align}

\subsection{Neutron Diffusion Method}

@@

\subsection{$S_N$ Neutron Transport Method}

@@

\subsection{Hybrid $S_N$-Diffusion Method}

>>>>

\section{Numerical Implementation of Neutronics Solvers on Python} \label{sec:implementation}

In Section \ref{sec:diffusion-correction}, I presented the theoretical basis for the hybrid
$S_N$-diffusion method with diffusion correction terms. This section presents numerical
implementation details of the neutron diffusion, $S_N$ neutron transport, and hybrid
$S_N$-diffusion solvers in the Python programming language for this work.

\subsection{Neutron Diffusion Method} \label{sec:python-diffusion}

@@

\subsection{$S_N$ Neutron Transport Method} \label{python-sn}

@@

\subsection{Hybrid $S_N$-Diffusion Method} \label{python-hybrid}

### 2024-08-11 hybrid.tex
msre chapter edits

Transport corrections to allow neutron diffusion methods to reproduce high-fidelity flux solutions
can be introduced by modifying the neutron diffusion equations.
In this work, I explored two options for applying
transport corrections to the neutron diffusion equations: applying
diffusion corrections or adding a drift correction term.
I performed preliminary investigations on 1-D test cases with both types of transport correction. I
eventually elected to use the drift correction term for the remaining test cases and reactor model
demonstrations due to superior numerical properties. I will make clear the reasons for this
selection through 1-D test case results in Chapter \ref{chap:hybrid-verification}.

\subsection{Diffusion Correction}

@@

For the hybrid $S_N$-diffusion method, we used a formulation that incorporates pointwise
corrections to the neutron diffusion flux solution from the $S_N$-derived flux solution as follows:
%
\begin{align}
  D^s_g(x) &= -J^{tr}_g(x)\bigg/\frac{d\phi^{tr}_g(x)}{dx}. \label{eq:svdc}
\end{align}
%
where $D^s_g$ is the \gls{SVDC}, and the $tr$ superscript denotes the transport-derived neutron
current and scalar flux solutions from the $S_N$ method. Transport corrections introduced through
$J_{tr}$ are scaled by the flux gradient. We assumed that it varies continuously and is
at least once differentiable except at dissimilar material interfaces. \glspl{SVDC} provide
pointwise corrections to closely match the diffusion flux solution to the $S_N$ flux solution.
By replacing $D_g$ with $D^s_g$, we are effectively adding the following transport correction term
%
\begin{gather}
  -\frac{\partial}{\partial x}(D^s_g-D_g)\frac{\partial\phi_g}{\partial x}
\end{gather}
to the neutron diffusion equations. Alternatively, we may define
$\partial D\equiv\left(D^s_g-D_g)\right)$. This alternative form shows that diffusion
correction applies a multiplicative closure that scales with the flux gradient.

@@

Unlike their approach, the \gls{SVDC} formulation in Eq.\ \ref{eq:svdc} allows for continuously
varying diffusion coefficients to reduce truncation error. In practice, the discretization order of
\gls{SVDC} variables in a numerical calculation would be the same as the discretization order of
the reference flux solution. This formulation introduces a minor change to the finite difference
implementation of the 2nd-order diffusion term to handle spatial derivatives of the
diffusion coefficient as $D^s_g$ is not uniform in space.

@@

\subsection{Drift Correction Term}

@@

As described in the previous section, transport correction formulations are commonly used in
iterative acceleration schemes or other hybrid methods to couple high-order neutron transport
calculations with low-order neutron diffusion calculations. However, the high-order calculations of
full reactor core models remain computationally intensive and limited in their applicability to
time-independent neutronics simulations on \gls{HPC} clusters.
In order to reduce the computational cost of the high-level $S_N$ calculation in a reactor
simulation, I propose reducing
the problem domain of the $S_N$ method to a \textit{correction region} containing the control rod
and its vicinity. Consequently, the hybrid $S_N$-diffusion method may retain accurate neutron flux
and current estimates around the control rod region from the $S_N$ method while making significant
computational cost savings by treating most of the reactor geometry with the neutron diffusion
method alone. Henceforth, I will refer to the $S_N$ calculation on the correction
region as the $S_N$ \textit{subproblem} or \textit{subsolver}. I define the full problem
domain and the correction region as $V_0$ and $V_1$, respectively, where
$V_1\subseteq V_0$. The iterative algorithm for the hybrid $S_N$-diffusion method is as follows:
%
\begin{enumerate}
  \item Start with an initial neutron diffusion calculation in $V_0$ using the standard neutron
    diffusion method.
  \item Pass the neutron diffusion current estimates along
    $\partial V_1$ to the $S_N$ subsolver to evaluate the boundary conditions for the $S_N$
    subproblem.
  \item With the $S_N$ subsolver, evaluate transport correction terms in $V_1$ using Eq.
    \ref{eq:svdc} or \ref{eq:drift}.
  \item Pass the transport correction terms to the neutron diffusion solver to apply corrections
    within $V_1$ while continuing to apply the standard neutron diffusion solver
    in the rest of $V_0$.
  \item Start the next iteration by running a neutron diffusion calculation with transport
    corrections in $V_1$.
  \item Repeat Steps 2-6 until convergence is reached by meeting desired convergence tolerance
    values.
\end{enumerate}
%
Figure \ref{fig:algorithm} shows a visual flowchart of the same hybrid method algorithm.

@@

The $S_N$ subsolver boundary conditions will generally cause deviations
in the flux and transport correction values since the $P_1$ approximation is accurate for up to
linear anisotropies in the flux. However, as I will show in Chapter \ref{chap:hybrid-verification},
the influence of boundary conditions on the accuracy of
the transport correction terms does not extend far from the boundary $\partial V_1$ in optically
thick media due to competing influences from the bulk scattering and absorption effects in $V_1$.
Results in Chapter \ref{chap:hybrid-verification} will show that transport correction values
($D^s_g$ or $\vec{D}_g$) generated with the $S_N$ subsolver within $V_1$ are accurate everywhere up
to a few centimeters away from $\partial V_1$.

@@

Here, I use Case 3b of the 1-D test cases to be covered in Chapter \ref{chap:hybrid-verification}
to illustrate the buffer region and the behavior of $\vec{D}_g$. Case 3b is a 1-D slab problem
consisting of control rod, air, fuel-graphite repeating lattice, and core vessel regions as shown
in Figure \ref{fig:3b-geometry}. The fuel regions are neutron-multiplying media containing
$^\text{235}$U fissile material. I took material compositions of each region from the \gls{MSRE}
reference design and the start-up fuel composition \cite{fratoni_molten_2020}.
Case 3b has reflecting boundary conditions on the left boundary and vacuum boundary conditions on
the right boundary.
%
\begin{figure}[htb!]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-1-drift}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-8-drift}
    \end{subfigure}
    \caption{The reference and hybrid drift ($\vec{D}_g$) distributions for group 1 and 8 calculated
      from $S_8$ and $S_8$-diffusion simulations. The correction region $V_1$ spans $x=0$ cm to
      $x=10$ cm.}
    \label{fig:3b-drift}
\end{figure}

>>>>

Transport corrections to allow neutron diffusion methods to reproduce high-fidelity flux solutions
can be introduced by modifying the neutron diffusion equations.
In this work, I explored two options for applying
transport corrections to the neutron diffusion equations: applying
diffusion corrections or adding a drift correction term.
I performed preliminary investigations on 1-D test cases with both types of transport correction. I
eventually elected to use the drift correction term for the remaining test cases and reactor model
demonstrations due to superior numerical properties. I will make clear the reasons for this
selection through 1-D test case results in Chapter \ref{chap:msre}.

\subsection{Diffusion Correction} \label{sec:diffusion-correction}

@@

For the hybrid $S_N$-diffusion method, we used a formulation that incorporates pointwise
corrections to the neutron diffusion flux solution from the $S_N$-derived flux solution as follows:
%
\begin{align}
  D^s_g(x) &= -J^{tr}_g(x)\bigg/\frac{d\phi^{tr}_g(x)}{dx}. \label{eq:svdc}
\end{align}
%
where $D^s_g$ is the diffusion correction parameter, and the $tr$ superscript denotes the
transport-derived neutron
current and scalar flux solutions from the $S_N$ method. Transport corrections introduced through
$J_{tr}$ are scaled by the flux gradient. We assumed that it varies continuously and is
at least once differentiable except at dissimilar material interfaces. $D^s_g$ provides
pointwise corrections to closely match the diffusion flux solution to the $S_N$ flux solution.
By replacing $D_g$ with $D^s_g$, we are effectively adding the following transport correction term
%
\begin{gather}
  -\frac{\partial}{\partial x}(D^s_g-D_g)\frac{\partial\phi_g}{\partial x}
\end{gather}
to the neutron diffusion equations. Alternatively, we may define
$\partial D\equiv\left(D^s_g-D_g)\right)$. This alternative form shows that diffusion
correction applies a multiplicative closure that scales with the flux gradient.

@@

Unlike their approach, my diffusion correction scheme (Eq. \ref{eq:svdc} allows for continuously
varying diffusion coefficients to reduce truncation error. In practice, the discretization order of
$D^s_g$ in a numerical calculation would be the same as the discretization order of
the reference flux solution. This formulation introduces a minor change to the finite difference
implementation of the 2nd-order diffusion term to handle spatial derivatives of the
diffusion coefficient as $D^s_g$ is not uniform in space.

@@

\subsection{Drift Correction Term} \label{sec:drift-correction}

@@

As described in the previous section, transport correction formulations are commonly used in
iterative acceleration schemes or other hybrid methods to couple high-order neutron transport
calculations with low-order neutron diffusion calculations. However, the high-order calculations of
full reactor core models remain computationally intensive and limited in their applicability to
time-independent neutronics simulations on \gls{HPC} clusters.
In order to reduce the computational cost of the high-level $S_N$ calculation in a reactor
simulation, I propose reducing the problem domain of the $S_N$ method to a
\textit{correction subregion} containing the control rod
and its vicinity. Consequently, the hybrid $S_N$-diffusion method may retain accurate neutron flux
and current estimates around the control rod region from the $S_N$ method while making significant
computational cost savings by treating most of the reactor geometry with the neutron diffusion
method alone. Henceforth, I will refer to the $S_N$ calculation on the correction
region as the $S_N$ \textit{subproblem} or \textit{subsolver}. I define the full problem
domain and the correction subregion as $V_0$ and $V_1$, respectively, where
$V_1\subseteq V_0$. The iterative algorithm for the hybrid $S_N$-diffusion method is as follows:
%
\begin{enumerate}
  \item Start with an initial neutron diffusion calculation in $V_0$ using the standard neutron
    diffusion method.
  \item Pass the neutron diffusion current estimates along
    $\partial V_1$ to the $S_N$ subsolver to evaluate the boundary conditions for the $S_N$
    subproblem.
  \item With the $S_N$ subsolver, evaluate transport correction terms in $V_1$ using Eq.
    \ref{eq:svdc} or \ref{eq:drift}.
  \item Pass the transport correction terms to the neutron diffusion solver to apply corrections
    within $V_1$ while continuing to apply the standard neutron diffusion solver
    in the rest of $V_0$.
  \item Start the next iteration by running a neutron diffusion calculation with transport
    corrections in $V_1$.
  \item Repeat Steps 2-6 until convergence is reached by meeting desired convergence tolerance
    values.
\end{enumerate}
%
Figure \ref{fig:algorithm} shows a visual flowchart of the same hybrid method algorithm.

@@

The $S_N$ subsolver boundary conditions will generally cause deviations
in the flux and transport correction values since the $P_1$ approximation is accurate for up to
linear anisotropies in the flux. However, as I will show in Chapter \ref{chap:msre},
the influence of boundary conditions on the accuracy of
the transport correction terms does not extend far from the boundary $\partial V_1$ in optically
thick media due to competing influences from the bulk scattering and absorption effects in $V_1$.
Results in Chapter \ref{chap:msre} will show that transport correction values
($D^s_g$ or $\vec{D}_g$) generated with the $S_N$ subsolver within $V_1$ are accurate everywhere up
to a few centimeters away from $\partial V_1$.

@@

Here, I use Case 3b of the 1-D test cases to be covered in Chapter \ref{chap:msre}
to illustrate the buffer region and the behavior of $\vec{D}_g$. Case 3b is a 1-D slab problem
consisting of control rod, air, fuel-graphite repeating lattice, and core vessel regions as shown
in Figure \ref{fig:3b-geometry}. The fuel regions are neutron-multiplying media containing
$^\text{235}$U fissile material. I took material compositions of each region from the \gls{MSRE}
reference design and the start-up fuel composition \cite{fratoni_molten_2020}.
Case 3b has reflecting boundary conditions on the left boundary and vacuum boundary conditions on
the right boundary.
%
\begin{figure}[htb!]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-1-drift}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{case-3b-group-8-drift}
    \end{subfigure}
    \caption{The reference and hybrid drift ($\vec{D}_g$) distributions for group 1 and 8 calculated
      from $S_8$ and $S_8$-diffusion simulations. The correction subregion $V_1$ spans $x=0$ cm to
      $x=10$ cm.}
    \label{fig:3b-drift}
\end{figure}
